TacticToe: Learning to Prove with Tactics

### 2018-03-19 tactictoe_jv.tex

  In this work, we develop in the ITP \holfour a procedure to select suitable
  tactics depending on the current proof state by learning
  from previous proofs. Backed by this machine-learned guidance, our prover
  \tactictoe relies on \emph{Monte
  Carlo tree search}~\cite{montecarlo} to find sequences of tactic applications
  leading to an ITP proof.
  To this end, we \emph{extract and record} tactic invocations from the ITP
  proofs (Section~\ref{sec:recording}) and
  \emph{build efficient machine learning classifiers} based on such training
  examples (Section~\ref{s:prediction}).  The learned data serves as
  guidance for the Monte Carlo tree search that explores different
  proof paths (Section~\ref{sec:proofsearch}). The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics (Section~\ref{sec:proofdisplay}). We evaluate the system on a large
  set of theorems originating
  from \holfour (Section~\ref{s:experiments}).

\subsection{The problem}
% How a user works with an ITP and how this gives rise to a machine learning problem.
\todo{RK: this is just a first draft -- feel free to edit lots!}
An ITP is a development environment for the construction of formal proofs in which it is possible, but not common, to write a proof as a sequence of applications of primitive inference rules.
The common approach, however, is to use and devise high-level tools and automation that abstract over useful ways of producing proof pieces.
A specific, prevalent instance of this approach is the use of \emph{tactics} for \emph{backward} or goal-directed proof.
Here, the ITP user operates on a proof state, initially the desired conjecture, and applies tactics that transform the proof state until it is solved (or until giving up).
Each tactic performs a sound transformation of the proof state: essentially, it promises that if the resulting proof state can be solved, then so can the starting one.
This gives rise to a machine learning problem: can we learn a mapping from a given proof state to the next tactic (or sequence of tactics) that will productively advance the proof towards a solution?

@@

\section{Case Study}
Since the proofs generated by \tactictoe are meant to be parts of a \holfour development, it interesting to compare their quality with the original human proofs.
The quality of a proof in \holfour can be measured in terms of length, readability, maintainability and verification speed.
We study these properties in three examples taken from our full-scale experiment.
We list the time needed by \holfour to check a proof in parentheses.

>>>>

  In this work, we develop in the ITP \holfour a procedure to select suitable
  tactics depending on the current proof state by learning
  from previous proofs. Backed by this machine-learned guidance, our prover
  \tactictoe relies on \emph{Monte
  Carlo tree search}~\cite{montecarlo} to find sequences of tactic applications
  leading to an ITP proof.

\subsection{The problem}
An ITP is a development environment for the construction of formal proofs in which it is possible, but not common, to write a proof as a sequence of applications of primitive inference rules.
The common approach, however, is to use and devise high-level tools and automation that abstract over useful ways of producing proof pieces.
A specific, prevalent instance of this approach is the use of \emph{tactics} for \emph{backward} or goal-directed proof.
Here, the ITP user operates on a proof state, initially the desired conjecture, and applies tactics that transform the proof state until it is solved (or until giving up).
Each tactic performs a sound transformation of the proof state: essentially, it promises that if the resulting proof state can be solved, then so can the starting one.
This gives rise to a machine learning problem: can we learn a mapping from a given proof state to the next tactic (or sequence of tactics) that will productively advance the proof towards a solution?

@@

\section{Case Study}\label{sec:case_study}
Since the proofs generated by \tactictoe are meant to be parts of a \holfour development, it interesting to compare their quality with the original human proofs.
The quality of a proof in \holfour can be measured in terms of length, readability, maintainability and verification speed.
We study these properties in three examples taken from our full-scale experiment.
We list the time needed by \holfour to check a proof in parentheses.
